Supporting Information:\\ NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-10-25 Manuscript/manuscriptPSsuppl.tex
Updates in the SI

\section{Headgroup response to the additional counterions in POPC:POPS (5:1) mixtures}\label{mixtureTOadditionalCIs}
To evaluate counterion binding in different simulation models against experimental data~\cite{roux90},
we plot the headgroup order parameters measured from POPC:POPS 5:1 mixture
as a function of different monovalent ions added to the buffer (Fig. \ref{PSresponseTONaCl}). 
Experimental order parameter data for POPC headgroup in the mixture is available as a function
of LiCl and KCl concentrations, while POPS headgroup order parameters are measured also
as a function of NaCl. Lithium interacts more strongly with PS headgroups than other monovalent 
ions~\cite{hauser83,hauser85,roux86,mattai89,roux90}, as also observed for PC headgroups~\cite{cevc90}. 
This is evident also in the changes of PS headgroup order parameters, which decrease with the addition of lithium 
but increase with the addition of sodium or potassium (Fig. \ref{PSresponseTONaCl}). 
POPC headgroup order parameters exhibit a clear decrease as a function of LiCl concentration
but only modest changes as a function of KCl concentration, indicating singificant 
Li$^+$ binding but only weak Na$^+$ binding to the mixture when interpreted using the
electrometer concept~\cite{akutsu81,altenbach84,seelig87}. In simulations with the
Berger model, the headgroup order parameter response of POPC
to the added NaCl is similar to the experiments of LiCl,
indicating overestimated binding affinity of sodium, in line with the results for PC bilayers \cite{catte16}.
Indeed, the sodium density profile shows a significant binding peak in the 
Berger model (Fig. \ref{CIdensPSOCmixt}). Potassium binding in the MacRog simulation
is significantly weaker  (Fig. \ref{CIdensPSOCmixt}) and the headgroup order parameter 
changes are also in better agreement with simulations (Fig. \ref{PSresponseTONaCl}).
\todo{Discussion about Lipid17 to be written when we have the density profiles.}
All the tested models overestimate the changes of POPS headgroup order parameters as
a function of monovalent ions (Fig. \ref{PSresponseTONaCl}), suggesting that
model development is necessary to interpret the PS headgroup-ion interactions 
from MD simulations.
\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
  \caption{\label{PSresponseTONaCl}
    Changes of the PC (left) and PS (right) headgroup order parameters as a function of
    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture at 298~K
    (except Berger simulations are (4:1) mixture at 310~K).
    The experimental data is from Ref. \citenum{roux90}.
    The values from counterion-only systems are set as a zero point of y-axis.
    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
    in PS headgroup (bottom, right), the y-axis is tranferred with the same value for both order parameters such that the lower order
    parameter value is at zero.
  }
\end{figure*}

\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/CIdensPSOCmixt.eps}
  \caption{  Counterion density distributions from PC:PS mixtures.
\label{CIdensPSOCmixt}
  }
  \todo{Lipid 17 is to be added when we have the density profiles.
  There is something wrong with the current plots in git: https://github.com/NMRLipids/MATCH/pull/40}
\end{figure}

@@

\begin{figure}[]
  \centering
  \includegraphics[width=17.0cm]{../Figs/OP_CHARMM_CaCl_POPC_NBFix.pdf}
  \caption{\label{OP_CHARMM_CaCl_POPC_NBFix}
    Headgroup order parameters from CHARMM36 simulations of POPC mixed with ions having
    the NBfix term employed for sodium \cite{venable13} ({\it left})
    and calcium \cite{kim16} ({\it right}) compared with the experimental data \cite{akutsu81,altenbach84}.
    Simulation files without ions are available at Ref. \citenum{nencineCHARMM36popcT310K},
    with the NBfix term in sodium at Ref. \citenum{CHARMM36withCaClNBfix}, 
    with the NBfix term in calcium at Ref. \citenum{CHARMM36withCaClNBfix} and
    without the NBfix in calcium at Ref. \citenum{charmmPOPC450mMCaClfiles}.
  }
\end{figure}

>>>>

\section{Headgroup response to the additional counterions in POPC:POPS (5:1) mixtures}\label{mixtureTOadditionalCIs}
To evaluate counterion binding in different simulation models against experimental data~\cite{roux90},
we plot the headgroup order parameters measured from POPC:POPS 5:1 mixture
as a function of different monovalent ions added to the buffer (Fig. \ref{PSresponseTONaCl}). 
Experimental order parameter data for POPC headgroup in the mixture is available as a function
of LiCl and KCl concentrations, while POPS headgroup order parameters are measured also
as a function of NaCl. Lithium interacts more strongly with PS headgroups than other monovalent 
ions~\cite{hauser83,hauser85,roux86,mattai89,roux90}, as also observed for PC headgroups~\cite{cevc90}. 
This is evident also in the changes of PS headgroup order parameters, which decrease with the addition of lithium 
but increase with the addition of sodium or potassium (Fig. \ref{PSresponseTONaCl}). 
POPC headgroup order parameters exhibit a clear decrease as a function of LiCl concentration
but only modest changes as a function of KCl concentration, indicating singificant 
Li$^+$ binding but only weak Na$^+$ binding to the mixture when interpreted using the
electrometer concept~\cite{akutsu81,altenbach84,seelig87}. In simulations with the
Berger model, the headgroup order parameter response of POPC
to the added NaCl is similar to the experiments of LiCl,
indicating overestimated binding affinity of sodium, in line with the results for PC bilayers \cite{catte16}.
Indeed, the sodium density profile shows a significant binding peak in the 
Berger model (Fig. \ref{CIdensPSOCmixt}). Potassium binding in the MacRog simulation
is significantly weaker  (Fig. \ref{CIdensPSOCmixt}) and the headgroup order parameter 
changes are also in better agreement with simulations (Fig. \ref{PSresponseTONaCl}).
\todo{Discussion about Lipid17 have be written based on issue https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/21}
All the tested models overestimate the changes of POPS headgroup order parameters as
a function of monovalent ions (Fig. \ref{PSresponseTONaCl}), suggesting that
model development is necessary to interpret the PS headgroup-ion interactions 
from MD simulations.
\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
  \caption{\label{PSresponseTONaCl}
    Changes of the PC (left) and PS (right) headgroup order parameters as a function of
    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture at 298~K
    (except Berger simulations are (4:1) mixture at 310~K).
    The experimental data is from Ref. \citenum{roux90}.
    The values from counterion-only systems are set as a zero point of y-axis.
    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
    in PS headgroup (bottom, right), the y-axis is tranferred with the same value for both order parameters such that the lower order
    parameter value is at zero.
  }
\end{figure*}

\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/CIdensPSOCmixt.eps}
  \caption{  Counterion density distributions from PC:PS mixtures.
\label{CIdensPSOCmixt}
  }
\end{figure}

@@

\begin{figure}[]
  \centering
  \includegraphics[width=17.0cm]{../Figs/OP_CHARMM_CaCl_POPC_NBFix.pdf}
  \caption{\label{OP_CHARMM_CaCl_POPC_NBFix}
    Headgroup order parameters from CHARMM36 simulations of POPC mixed with ions having
    the NBfix term employed for sodium \cite{venable13} ({\it left})
    and calcium \cite{kim16} ({\it right}) compared with the experimental data \cite{akutsu81,altenbach84}
    and simulations without NBfix for the calcium.
    Simulation files without ions are available at Ref. \citenum{nencineCHARMM36popcT310K},
    with the NBfix term in sodium at Ref. \citenum{CHARMM36withCaClNBfix}, 
    with the NBfix term in calcium at Ref. \citenum{CHARMM36withCaClNBfix} and
    without the NBfix in calcium at Ref. \citenum{charmmPOPC450mMCaClfiles}.
  }
\end{figure}
